Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-12 paper.tex
First version of GeL data analysis

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first convert the X VCF files (YTiB) for chromosome 2
into the intermediate columnar format. This task was 
distributed over a cluster of N workers, using a 
total of X CPU hours and an elapsed wall-clock time 
of X hours. When then converted the intermediate 
column format to Zarr using 
SOMETHING which took X hours using [RESOURCES]. 
This produced a dataset with N arrays, consuming a 
total of Y TiB of storage over X directories and 
Y files. The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary of the top N vcfzarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.0G & 26 & 24.1\\
call\_GQ &  int16 & 171.0G & 13 & 22.2 \\
call\_DP &  int16 & 141.0G & 20 & 18.8 \\
call\_FT &  string & 58.0G & 160 & 7.3 \\
call\_PL &  int16 & 51.0G & 140 & 6.8 \\
call\_GQX &  int16 & 12.0G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.6G  & 630 & 0.48\\
call\_genotype\_phased & bool & 3.6G  & 1700 & 0.48 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size dominated by a few columns, with call\_AD and call\_GQ consuming 
almost half of total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
due to their inherent noisiness~\citep{lin2020sparse}. Substantial
reductions in space can be achieved, however, if we examine the 
distributions of these fields and truncate some outliers.
Figure~SX shows the histogram of the XX call\_AD values, and we can
see that < 0.0X\% of the values are $>127$. Truncating and 
re-encoding the call\_AD array as an int8 results in a total storage of 
XX. A substantial advantage of Zarr is that such per-field optimisations
can be made after conversion, without affecting the rest of the dataset.
As discussed in Methods, judicious use of additional filters
and lossy quantizing can result substantial space savings.

Calculating the histogram of call\_AD values at this 
scale is not a trivial task.
Taking advantage of the rich software ecosystem growing around Zarr
(see Discussion) we used xarray~\citep{hoyer2017xarray}
and Dask~\citep{rocklin2015dask} to transparently distribute the 
computation over a cluster of X workers. The elapsed time 
was X minutes, and requires X lines of Python in a 
Jupyter notebook~\citep{kluyver2016jupyter}.

>>>>

As described in the Methods, conversion to Zarr using 
\texttt{vcf2zarr} is a two-step 
process to limit memory usage and ensure reliability. We 
first converted the 28 VCF files (3.51 TiB) for chromosome 20
into the intermediate columnar format (ICF). This task was 
split into 3899 partitions, and distributed using the Genomics England
HPC cluster. It required about 50 days of CPU, with an elapsed
wall-clock time of about 2 hours. The ICF representation uses a total
of XTiB over XX files and YY directories. We then encoded this intermediate
columnar format to Zarr using 8 cores and XX GB of RAM in about 12 hours.
This produced a dataset with 43 arrays, consuming a 
total of Y TiB of storage over X directories and 
Y files. The top fields in terms 
of storage are detailed in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary for a selection of the largest vcfzarr columns produced for 
Genomics England aggV2 VCFs on chromosome 20 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes.
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
call\_AD &  int16 & 179.7G & 26 & 24.0\\
call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
call\_DP &  int16 & 141.8G & 16 & 18.9 \\
call\_DPF& int16  & 115.1G & 20 & 15.3\\
call\_FT &  string & 58.5G & 160 & 7.8 \\
call\_PL &  int16 & 51.4G & 140 & 6.9 \\
call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
call\_genotype & int8 & 6.1G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
%         6.85333333,  1.61066667,  0.81733333,  0.488     ])

% /call_AD                     int16    179.67 GiB  4.53 TiB       26 125768  37.74 MiB     1.46 MiB            (15912293, 78195, 2)  (10000, 1000, 2)  
% /call_GQ                     int16    171.79 GiB  2.26 TiB       13 125768  18.87 MiB     1.4 MiB             (15912293, 78195)     (10000, 1000) 
% /call_DP                     int16    141.76 GiB  2.26 TiB       16 125768  18.87 MiB     1.15 MiB            (15912293, 78195)     (10000, 1000)
% /call_DPF                    int16    115.09 GiB  2.26 TiB       20 125768  18.87 MiB     959.56 KiB          (15912293, 78195)     (10000, 1000)
% /call_FT                     object   58.5 GiB    9.05 TiB      160 125768  75.48 MiB     487.71 KiB          (15912293, 78195)     (10000, 1000)
% /call_PL                     int16    51.4 GiB    6.79 TiB      140 125768  56.61 MiB     428.51 KiB          (15912293, 78195, 3)  (10000, 1000, 3)
% /call_GQX                    int16    12.08 GiB   2.26 TiB      190 125768  18.87 MiB     100.73 KiB          (15912293, 78195)     (10000, 1000) 
% /call_genotype               int8     6.13 GiB    2.26 TiB      380 125768  18.87 MiB     51.11 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
% /call_genotype_mask          bool     3.66 GiB    2.26 TiB      630 125768  18.87 MiB     30.53 KiB           (15912293, 78195, 2)  (10000, 1000, 2)  
% /call_genotype_phased        bool     692.51 MiB  1.13 TiB     1700 125768  9.43 MiB      5.64 KiB            (15912293, 78195)     (10000, 1000) 
% /call_PS                     int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000) 
% /call_ADR                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
% /call_ADF                    int8     102.2 MiB   1.13 TiB    12000 125768  9.43 MiB      852 bytes           (15912293, 78195)     (10000, 1000)
% /variant_OLD_MULTIALLELIC    object   31.63 MiB   121.4 MiB       3.8 1592  78.09 KiB     20.34 KiB           (15912293,)           (10000,) 
% /variant_quality             float32  22.6 MiB    60.7 MiB        2.7 1592  39.04 KiB     14.53 KiB           (15912293,)           (10000,) 
% /variant_allele              object   21.52 MiB   242.8 MiB      11 1592  156.17 KiB    13.84 KiB           (15912293, 2)         (10000, 2) 
% /variant_position            int32    12.83 MiB   60.7 MiB        4.7 1592  39.04 KiB     8.25 KiB            (15912293,)           (10000,) 
% /variant_ABratio             float32  9.62 MiB    60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
% /variant_AN                  int32    9.6 MiB     60.7 MiB        6.3 1592  39.04 KiB     6.18 KiB            (15912293,)           (10000,) 
% /variant_AC                  int32    9.22 MiB    60.7 MiB        6.6 1592  39.04 KiB     5.93 KiB            (15912293,)           (10000,) 
% /variant_AC_Het              int32    8.98 MiB    60.7 MiB        6.8 1592  39.04 KiB     5.78 KiB            (15912293,)           (10000,)

% Approx total = 179.67 + 171.79 + 141.76 + 115.09 + 58.5 + 51.4 + 12.08 + 6.13 + 3.66 
% 740, round up to 750
% [179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 , 6.13 , 3.66 ]
% >>> import numpy as np
% >>> a = np.array([179.67 , 171.79 , 141.76 , 115.09 , 58.5 , 51.4 , 12.08 ,
% 6.13 , 3.66 ]  )
% >>> a
% array([179.67, 171.79, 141.76, 115.09,  58.5 ,  51.4 ,  12.08,   6.13,
%          3.66])
% >>> a / 750 * 100
% array([23.956     , 22.90533333, 18.90133333, 15.34533333,  7.8       ,
%         6.85333333,  1.61066667,  0.81733333,  0.488     ])

Table~\ref{tab-genomics-england-data} shows that the dataset storage
size is dominated by a few columns with the top four
(call\_AD, call\_GQ, call\_DP and call\_DPF) accounting for over 
80\% of the total. These fields are much less compressible
than genotype data (which uses $<1\%$ of the total space here)
because of their inherent noisiness~\citep{lin2020sparse}. By 
default the \texttt{bio2zarr} conversion is lossless, and integers
of the narrowest type required to represent all of the data are
used. In this case, this has resulted in 16 bit integers being used 
for these top four fields. Examining the distribution of the depth
fields, (call\_AD, call\_DP, call\_DPF), we can 
we can see that a tiny fraction of the values stored
are greater than 127 (call\_AD: 0.1\%, call\_DP: xx\%, call\_DPF: x\%; see 
Fig SX for histograms).
One might then reasonably truncate these fields to 8 bit integers.
After converting to 8 bit, call\_AD is reduced by 
x\% (XX GiB) and call\_DP reduced by x\% (XX GiB)
and call\_DPF by x\% (XX GiB). 
Similarly, the vast majority of call\_GQ values (xx\%) are less than 
128 (corresponding to a conditional probability that the genotype call is 
wrong of $10^{-128}$ [TODO CHECK THIS]), and can also reasonably
be truncated. Here, this results in a reduction of x\% (XX GiB).

% and can reasonably be truncated (a 
% much information would be lost if we used 8 bit
% Substantial
% reductions in space can be achieved, however, if we examine the 
% distributions of these fields and truncate some outliers.
% Figure~SX shows the histogram of the XX call\_AD values, and we can
% see that < 0.0X\% of the values are $>127$. Truncating and 
% re-encoding the call\_AD array as an int8 results in a total storage of 
% XX. A substantial advantage of Zarr is that such per-field optimisations
% can be made after conversion, without affecting the rest of the dataset.
% As discussed in Methods, judicious use of additional filters
% and lossy quantizing can result substantial space savings.

Calculating the histogram of per call values at this 
scale is not a trivial task.
Taking advantage of the rich software ecosystem growing around Zarr
(see Discussion) we used xarray~\citep{hoyer2017xarray}
and Dask~\citep{rocklin2015dask} to transparently distribute the 
computations over a cluster of X workers. 
[ADD DETAILS]
The elapsed time was X minutes, and requires X lines of Python in a 
Jupyter notebook~\citep{kluyver2016jupyter}.
